---
title: "04 Hierarchical modeling"
author: "Shravan Vasishth"
date: '```r format(Sys.Date(), "%B %d, %Y")```'
output:
  beamer_presentation:
    theme: "Boadilla"
    colortheme: "dove"
    fonttheme: "structurebold"
header-includes:
   - \usepackage{esint}
   - \usepackage{mathtools}
   - \makeatletter
   - \newcommand{\explain}[2]{\underset{\mathclap{\overset{\uparrow}{#2}}}{#1}}
   - \newcommand{\explainup}[2]{\overset{\mathclap{\underset{\downarrow}{#2}}}{#1}}
   - \makeatother
citation_package: biblatex
biblatexoptions: 
  - "backend=biber, style=apa"
bibliography:  bayes.bib
link-citations: yes
---



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(rstan)
library(brms)
library(lme4)
```

# Example 1: Reading time differences in subject vs object relatives in English

We begin with a relatively simple question from the psycholinguistics literature: are subject relatives easier to process than object relatives? The data come from Experiment 1 in a paper by @grodner.

# Example 1: Reading time differences in subject vs object relatives in English
## Scientific question: Is there a subject relative advantage in reading?


(1a) The *reporter* who the photographer *sent* to the editor was hoping for a good story. (object gap)

(1b) The *reporter* who *sent* the photographer to the editor was hoping for a good story. (subject gap)


# Example 1: Reading time differences in subject vs object relatives in English
## Load data and reformat

```{r loaddata,echo=TRUE}
library(bcogsci)
data("df_gg05_rc")
```



# Example 1: Reading time differences in subject vs object relatives in English
## Experiment design: Latin square and crossed subject and items

Two important properties of these data are worth noticing. 

  - Latin square design
  - Crossed subjects and items

# Example 1: Reading time differences in subject vs object relatives in English
## Latin-square design

First, the design is the classic repeated measure Latin square set-up. To see what this means, 
first look at the number of subjects and items, and the number of rows in the data frame:

```{r}
length(unique(df_gg05_rc$subj))
length(unique(df_gg05_rc$item))
dim(df_gg05_rc)[1]
```

There are 42 subjects and 16 items. There are $42\times 16 = 672$ rows in the data frame. 

# Example 1: Reading time differences in subject vs object relatives in English
## Latin-square design

Notice also that each subject sees exactly eight object gap and eight subject gap sentences:

```{r}
head(xtabs(~subj+condition,df_gg05_rc),n=4)
```

# Example 1: Reading time differences in subject vs object relatives in English
## Latin-square design

\begin{table}[!htbp]
\caption{The Latin-square design in repeated measures experiments.}
\begin{center}
\begin{tabular}{ccc}
item id & group 1 & group 2\\
1       & objgap       & subjgap \\
2       & subjgap       & objgap \\
3       & objgap       & subjgap \\
4       & subjgap       & objgap \\
\vdots  & \vdots  & \vdots \\
16      & subjgap       & objgap \\
\end{tabular}
\end{center}
\label{tab:latinsq}
\end{table}

# Example 1: Reading time differences in subject vs object relatives in English
## Latin-square design: Ensuring balance

```{r}
sample(rep(c("order1","order2"),11))
```

Latin square designs are used in planned experiments because they are optimal in several ways.

# Example 1: Reading time differences in subject vs object relatives in English
## Latin-square design: Generating fake data

```{r}
library(MASS)
nitem <- 16
nsubj <- 42
## prepare data frame for two condition in a latin square design:
g1<-data.frame(item=1:nitem,
                 cond=rep(c("objgap","subjgap"),nitem/2))
g2<-data.frame(item=1:nitem,
                 cond=rep(c("subjgap","objgap"),nitem/2))
```

# Example 1: Reading time differences in subject vs object relatives in English
## Latin-square design: Generating fake data


```{r}
## assemble data frame in long format:
gp1<-g1[rep(seq_len(nrow(g1)),
              nsubj/2),]
gp2<-g2[rep(seq_len(nrow(g2)),
              nsubj/2),]

fakedat<-rbind(gp1,gp2)
dim(fakedat) ## sanity check
fakedat$subj<-rep(1:nsubj,each=nitem) ## add subjects
fakedat<-fakedat[,c(3,1,2)]  
fakedat$so<-ifelse(fakedat$cond=="objgap",1,-1)
```

# Example 1: Reading time differences in subject vs object relatives in English
## Latin-square design: Generating fake data


For example, subject 1 sees the following conditions and items:

```{r}
head(fakedat,n=16)
```

We will need this code later for fake data simulation.


# Example 1: Reading time differences in subject vs object relatives in English
## Fully crossed subjects and items

In the data, because of the Latin square design, each subject sees exactly one item in one of the two conditions:

```{r}
xtabs(~subj+item,df_gg05_rc)
```

# Example 1: Reading time differences in subject vs object relatives in English
## The implied generative model

The above design implies a particular statistical model that takes us beyond the linear model.

To remind you, a simple linear model of the above data would be:

\begin{equation}
y \sim Normal(\alpha + \beta * so, \sigma) 
\end{equation}

Here, object gaps are coded +1, subject gaps -1. See @SchadEtAlcontrasts for an explanation of contrast coding.

```{r}
df_gg05_rc$so<-ifelse(df_gg05_rc$condition=="objgap",1,-1)
```


# Example 1: Reading time differences in subject vs object relatives in English
## The implied generative model

As figure \ref{fig:ggrtdistrn} shows, a Normal likelihood doesn't seem well motivated, so we will use the log-normal.

```{r echo=FALSE,fig.height=3,fig.cap="\\label{fig:ggrtdistrn}Distribution of reading times in the Grodner and Gibson Experiment 1 data, at the critical region."}
plot(density(df_gg05_rc$RT),main="Grodner and Gibson Expt 1",xlab="RTs (ms)")
```

# Example 1: Reading time differences in subject vs object relatives in English
## The implied generative model

\begin{equation}
y \sim LogNormal(\alpha + \beta * so, \sigma) 
\end{equation}


# Example 1: Reading time differences in subject vs object relatives in English
## Between subject variability in mean reading time


```{r echo=FALSE,fig.height=3,fig.cap="\\label{fig:subjint}Between subject variability in mean reading times."}
hist(with(df_gg05_rc,tapply(RT,subj,mean)),
     main="Between subject variability",
     xlab="mean RTs",freq=FALSE)
```


# Example 1: Reading time differences in subject vs object relatives in English
## Between subject variability in mean reading time

In the linear model, we can express the assumption that the grand mean intercept $\alpha$ needs an adjustment by subject. 

\textbf{Here, subj[n]=subject id in n-th row of data}.

\begin{equation}
y_n \sim LogNormal(\alpha + u_{subj[n],1]} + \beta * so_n, \sigma)
\end{equation}

where we now have two sources of variance:

- within subject: $\sigma$ 
- between subject variance in mean reading times:  $u_{subj[n],1} \sim Normal(0,\sigma_{u1})$ 


# Example 1: Reading time differences in subject vs object relatives in English
## Between item variability in mean reading time

```{r echo=FALSE,fig.height=3,fig.cap="\\label{fig:itemint}Between item variability in mean reading times."}
hist(with(df_gg05_rc,tapply(RT,item,mean)),
     main="Between item variability",
     xlab="mean RTs",freq=FALSE)
```

# Example 1: Reading time differences in subject vs object relatives in English
## Between item variability in mean reading time


For items ranging from $j=1,\dots,16$, we can add this assumption to the model:

\begin{equation}
y_{n} \sim LogNormal(\alpha + u_{subj[n],1} + w_{item[n],1} + \beta * so_{n},\sigma)
\end{equation}

where there are now three variance components:

- $\sigma$
- $u_{subj[n],1} \sim Normal(0,\sigma_{subj[n],1})$
- between item variability in mean reading time, $w_{item[n],1} \sim Normal(0,\sigma_{w1})$

This model is called a *varying intercepts model* with crossed varying intercepts for subjects and for items.

# Example 1: Reading time differences in subject vs object relatives in English
## Between subject and between item variability in objgap cost


```{r echo=FALSE,fig.cap="\\label{fig:subjitemslope}Between subject and item variability in object gap vs subject gap reading times."}
op<-par(mfrow=c(1,2),pty="s")
meanssubj<-with(df_gg05_rc,
                tapply(RT,IND=list(subj,condition),mean))
diff<-meanssubj[,2]-meanssubj[,1]

hist(diff,
     main="Between subject variability",xlab="mean objgap cost",freq=FALSE,xlim=c(-800,200),
     ylim=c(0,0.005))

meansitem<-with(df_gg05_rc,
                tapply(RT,IND=list(item,condition),mean))
diff<-meansitem[,2]-meansitem[,1]

hist(diff,
     main="Between item variability",xlab="mean objgap cost",freq=FALSE,xlim=c(-800,200),ylim=c(0,0.005))
```

# Example 1: Reading time differences in subject vs object relatives in English
## Between subject and between item variability in objgap cost

We can incorporate this assumption into the model by adding adjustments to the $\beta$ parameter:

\begin{equation}
y_{n} \sim LogNormal(\alpha + u_{subj[n],1} + w_{item[n],1} + (\beta + u_{subj[n],2} + w_{item[n],1})* so_{n},\sigma) 
\end{equation}

where 

  - $\sigma$ 
  - $u_{subj[n],1} \sim Normal(0,\sigma_{u1})$
  - $u_{subj[n],2} \sim Normal(0,\sigma_{u2})$
  - $w_{item[n],1} \sim Normal(0,\sigma_{w1})$
  - $w_{item[n],2} \sim Normal(0,\sigma_{w2})$

This is called the *varying intercepts and slopes* model with *no correlation* between the intercepts and slopes.

# Example 1: Reading time differences in subject vs object relatives in English
## The "maximal" model \label{maximal}

  - There is one detail still missing in the model: the adjustments to the intercept and slope are correlated for subjects, and also for items. 
  - In other words, we have a bivariate distribution for the subject and item random effects:

# Example 1: Reading time differences in subject vs object relatives in English
## The "maximal" model 


\begin{equation}
y_{n} \sim LogNormal(\alpha + u_{subj[n],1} + w_{item[n],1} + (\beta + u_{subj[n],2} + w_{item[n],2}) * so_{n}, \sigma)
\end{equation}

where we have variance components: $\sigma$ and 

\begin{equation}\label{eq:covmat}
\Sigma _u
=
\begin{pmatrix}
\sigma _{u1}^2  & \rho _{u}\sigma _{u1}\sigma _{u2}\\
\rho _{u}\sigma _{u1}\sigma _{u2}    & \sigma _{u2}^2\\
\end{pmatrix}
\quad 
\Sigma _w
=
\begin{pmatrix}
\sigma _{w1}^2  & \rho _{w}\sigma _{w1}\sigma _{w2}\\
\rho _{w}\sigma _{w1}\sigma _{w2}    & \sigma _{w2}^2\\
\end{pmatrix}
\end{equation}

\begin{equation}\label{eq:jointpriordist1}
\begin{pmatrix}
  u_1 \\ 
  u_2 \\
\end{pmatrix}
\sim 
\mathcal{N} \left(
\begin{pmatrix}
  0 \\
  0 \\
\end{pmatrix},
\Sigma_{u}
\right),
\quad
\begin{pmatrix}
  w_1 \\ 
  w_2 \\
\end{pmatrix}
\sim 
\mathcal{N}\left(
\begin{pmatrix}
  0 \\
  0 \\
\end{pmatrix},
\Sigma_{w}
\right)
\end{equation}

# Example 1: Reading time differences in subject vs object relatives in English
## The maximal model 

This is a varying intercepts and slopes model with fully specified  variance-covariance matrices for the subject and item random effects. It is sometimes called the **maximal model** [@barr2013].

# Example 1: Reading time differences in subject vs object relatives in English
## Implementing the model

The above model is simple to implement in the Bayesian framework.

# Example 1: Reading time differences in subject vs object relatives in English
## Specify and visualize priors

Define some priors first:

\begin{enumerate}
\item $\alpha \sim Normal(0,10)$
\item $\beta \sim Normal(0,1)$
\item Residual standard deviation: $\sigma \sim Normal_{+}(0,1)$
\item All other standard deviations: $\sigma \sim Normal_{+}(0,1)$
\item Correlation matrix: $\rho \sim LKJcorr(2)$. 
\end{enumerate}


# Example 1: Reading time differences in subject vs object relatives in English
## The LKJ prior on the correlation matrix

  - In this model, we assume that the vector 
$\mathbf{u}=\langle u_1, u_2 \rangle$
comes from a bivariate normal distribution with a variance-covariance matrix $\boldsymbol{\Sigma_u}$.
  - This matrix has the variances of the adjustment to the intercept and to the slope respectively along the diagonal, and the covariance on the off-diagonals. 

# Example 1: Reading time differences in subject vs object relatives in English
## The LKJ prior on the correlation matrix

  - Recall that the covariance $Cov(X,Y)$ between two variables $X$ and $Y$ is defined as the product of their correlation $\rho$ and their standard
deviations $\sigma_X$ and $\sigma_Y$, such that, $Cov(X,Y) = \rho \sigma_X \sigma_Y$.

\begin{equation}
\boldsymbol{\Sigma_u} = 
{\begin{pmatrix} 
\sigma_{u_1}^2 & \rho_u \sigma_{u_1} \sigma_{u_2} \\ 
\rho_u \sigma_{u_1} \sigma_{u_2} & \sigma_{u_2}^2
\end{pmatrix}}
\end{equation}


# Example 1: Reading time differences in subject vs object relatives in English
## The LKJ prior on the correlation matrix

The covariance matrix can be decomposed into a vector of standard deviations and a correlation matrix. The correlation matrix looks like this:

\begin{equation}
{\begin{pmatrix} 
1 & \rho_u  \\ 
\rho_u  & 1
\end{pmatrix}}
\end{equation}

# Example 1: Reading time differences in subject vs object relatives in English
## The LKJ prior on the correlation matrix

In Stan, we write a matrix that has 0's on the off-diagonals as:

\begin{equation}
diag\_matrix(\sigma_{u_1},\sigma_{u_2}) = 
\begin{pmatrix} 
\sigma_{u_1} & 0 \\ 
0  & \sigma_{u_2}
\end{pmatrix}
\end{equation}

This means that we can decompose the covariance matrix into three parts:

\begin{equation}
\begin{aligned}
\boldsymbol{\Sigma_u} &= diag\_matrix(\sigma_{u_1},\sigma_{u_2}) \cdot \boldsymbol{\rho_u} \cdot diag\_matrix(\sigma_{u_1},\sigma_{u_2})\\
&=
{\begin{pmatrix} 
\sigma_{u_1} & 0 \\ 
0  & \sigma_{u_2}
\end{pmatrix}}
{\begin{pmatrix} 
1 & \rho_u  \\ 
\rho_u  & 1
\end{pmatrix}}
{\begin{pmatrix} 
\sigma_{u_1} & 0 \\ 
0  & \sigma_{u_2}
\end{pmatrix}}
\end{aligned}
\end{equation}

# Example 1: Reading time differences in subject vs object relatives in English
## The LKJ prior on the correlation matrix

So we need priors for the $\sigma_u$'s and for $\rho_u$:

  - The basic idea of the  LKJ prior is that its parameter (usually called *eta*, $\eta$, here it has value $2$) increases, the prior increasingly concentrates around the unit correlation matrix (i.e., favors smaller correlation: ones in the
diagonals and values close to zero in the lower and upper triangles). 
  - At $\eta = 1$, the LKJ correlation distribution is uninformative (similar to $Beta(1,1)$), at $\eta < 1$, it favors extreme correlations  (similar to $Beta(a<1,b<1)$).

# Example 1: Reading time differences in subject vs object relatives in English
## Visualize the priors

As always, it is a good idea to visualize these priors. See Figure \ref{fig:priorsgg}.

```{r results="hide",echo=FALSE,message=FALSE,warning=FALSE}
priors_alpha <- c(0,10)
priors_beta <- c(0,1)
priors_sigma_e <- c(0,1)
priors_sigma_u <- c(0,1)
priors_sigma_w <- c(0,1)

## code for visualizing lkj priors:
fake_data <- list(x = rnorm(30,0,1),
                  N = 30, R = 2) 
stancode <- "
data {
  int<lower=0> N; 
  real x[N]; 
  int R;
  }
parameters {
  real mu;
  real<lower=0> sigma;
}
model {
  x ~ normal(mu,sigma);  
}
generated quantities {
  corr_matrix[R] LKJ05;
  corr_matrix[R] LKJ1;
  corr_matrix[R] LKJ2;
  corr_matrix[R] LKJ4;
  LKJ05 = lkj_corr_rng(R,.5);
  LKJ1 = lkj_corr_rng(R,1);
  LKJ2 = lkj_corr_rng(R,2);
  LKJ4 = lkj_corr_rng(R,4);
}
"

fitfake <- stan(model_code = stancode, pars = c("LKJ05","LKJ1","LKJ2","LKJ4"),
                data = fake_data, chains = 4, 
                iter = 2000)

corrs<-extract(fitfake,pars=c("LKJ05[1,2]","LKJ1[1,2]","LKJ2[1,2]","LKJ4[1,2]"))
```

# Example 1: Reading time differences in subject vs object relatives in English
## Visualize the priors

```{r visualizepriors,echo=FALSE,fig.height=3,fig.cap="\\label{fig:priorsgg}Priors for the Godner and Gibson data.",echo=FALSE}
op<-par(mfrow=c(2,3),pty="s")
par(oma = rep(0, 4), mar = c(2.7, 2.7, 0.1, 0.1), mgp = c(1.7, 0.4, 0))
b<-seq(-priors_alpha[2]*2,priors_alpha[2]*2,by=0.01)
plot(b,dnorm(b,mean=priors_beta[1],sd=priors_beta[2]),type="l",ylab="density", 
     xlab=expression(alpha),ylim=c(0, 0.5))
plot(b,dnorm(b,mean=priors_beta[1],sd=priors_beta[2]),type="l",ylab="density",
     xlab=expression(beta),ylim=c(0, 0.5))
sig<-seq(0,priors_sigma_e[2]*3,by=0.01)
plot(sig,dnorm(sig,mean=priors_sigma_e[1],sd=priors_sigma_e[2]),type="l",ylab="density",
     xlab=expression(sigma[e]))
plot(sig,dnorm(sig,mean=priors_sigma_u[1],sd=priors_sigma_u[2]),type="l",ylab="density",
     xlab=expression(sigma[u[0]]))
plot(sig,dnorm(sig,mean=priors_sigma_u[1],sd=priors_sigma_u[2]),type="l",ylab="density",
     xlab=expression(sigma[w[0,1]]))
plot(density(corrs[[3]],bw=0.15),ylab="density",xlab=expression(rho),xlim=c(-1,1),main="")
```

# Example 1: Reading time differences in subject vs object relatives in English
## Fit the model using brms

```{r message=FALSE,warning=FALSE}
priors <- c(set_prior("normal(0, 10)", 
                      class = "Intercept"),
                      set_prior("normal(0, 1)", 
                                class = "b", 
                                coef = "so"),
                      set_prior("normal(0, 1)", 
                                class = "sd"),
                      set_prior("normal(0, 1)", 
                                class = "sigma"),
                      set_prior("lkj(2)", 
                                class = "cor"))
```

# Example 1: Reading time differences in subject vs object relatives in English
## Fit the model using brms


```{r results="hide",cache=TRUE,message=FALSE}
m_gg<-brm(RT~so + (1+so|subj) + (1+so|item),
          df_gg05_rc,family=lognormal(),
    prior=priors)
```

# Example 1: Reading time differences in subject vs object relatives in English
## Fit the model using brms

```{r message=FALSE,warning=FALSE,echo=FALSE,message=FALSE,fig.height=5}
stanplot(m_gg,type="hist")
```

# Example 1: Reading time differences in subject vs object relatives in English
## Fit the model using brms

Look at the posterior distributions of the parameters on the log ms scale (for the coefficients and standard deviations). Notice that

  - The object relative takes longer to read than the subject relative, as predicted. We know this because the parameter b_so is positive.
  - The largest sources of variance are the subject intercepts, slopes, and the residual standard deviation. Look at the sd_subject parameters, and sigma.
  - The by-item variance components are relatively small. Look at the sd_item parameters.
  - The correlations have very wide uncertainty---the prior is dominating in determining the posteriors as there isn't that much data to obtain accurate estimates of these parameters. Look at the cor parameters.

# Example 1: Reading time differences in subject vs object relatives in English
## Examine by subject random effects visually

First, extract the posterior samples of the parameters that we will need to compute individual differences.

```{r message=FALSE, warning=FALSE}
library(bayesplot)
```

```{r message=FALSE, warning=FALSE}
postgg<-posterior_samples(m_gg)
## extract variances:
alpha<-postgg$b_Intercept
beta<-postgg$b_so
cor<-posterior_samples(m_gg,"^cor")
sd<-posterior_samples(m_gg,"^sd")
sigma<-posterior_samples(m_gg,"sigma")
```

# Example 1: Reading time differences in subject vs object relatives in English
## Examine by subject random effects visually

```{r echo=FALSE,message=FALSE,warning=FALSE}
## item random effects won't be used below
item_re<-posterior_samples(m_gg,"^r_item")
subj_re<-posterior_samples(m_gg,"^r_subj")
```

# Example 1: Reading time differences in subject vs object relatives in English
## By subject slope adjustments


```{r echo=FALSE,fig.height=3,fig.cap="\\label{fig:ggsubjslope}Variability in subject slope adjustments in the Grodner and Gibson data."}
subjslope<-subj_re[,(1:42)+42]
colnames(subjslope)<-c(paste("u1,",1:42,sep=""))
slopemns <- colMeans(subjslope)
subjslope<-subjslope[,order(slopemns)]
mcmc_areas(subjslope)
```

# Example 1: Reading time differences in subject vs object relatives in English
## By subject slope adjustments


```{r echo=FALSE,warning=FALSE,fig.height=3,fig.cap="\\label{fig:ggsubjsloperidges}Variability in subject slope adjustments in the Grodner and Gibson data."}
mcmc_areas_ridges(subjslope)
```


# Example 1: Reading time differences in subject vs object relatives in English
## By subject slope adjustments

```{r echo=FALSE,warning=FALSE,fig.height=3,fig.cap="\\label{fig:ggpostcorr}Posterior distributions of subject varying intercept and slope correlation parameter in the Grodner and Gibson data."}
stanplot(m_gg,type="hist",pars="cor_subj__Intercept__so")
```

# Example 1: Reading time differences in subject vs object relatives in English
## Examine mean and individual differences on the raw ms scale 

It is useful to see the effects on the raw ms scale. The log ms scale is difficult to interpret. 

# Example 1: Reading time differences in subject vs object relatives in English
## Mean difference

```{r}
exp(mean(alpha)+mean(beta))-exp(mean(alpha)-mean(beta))
meandiff<- exp(alpha + beta) - exp(alpha - beta)
mean(meandiff)
round(quantile(meandiff,prob=c(0.025,0.975)),0)
```

# Example 1: Reading time differences in subject vs object relatives in English
## Mean difference


```{r echo=FALSE,fig.height=3,fig.cap="\\label{fig:ggmeandiff}Mean OR processing cost effect in the Grodner and Gibson data."}
hist(meandiff,freq=FALSE,
     main="Mean OR vs SR processing cost",
     xlab=expression(exp(alpha + beta)- exp(alpha - beta)))
```

# Example 1: Reading time differences in subject vs object relatives in English
## Individual effects of OR processing cost


```{r echo=FALSE,fig.height=3,fig.cap="\\label{fig:ggsubjeffect}Variability in subject OR processing cost effect in the Grodner and Gibson data."}
subjdiff<-matrix(rep(NA,42*4000),nrow=42)
for(i in 1:42){
subjdiff[i,]<-exp(alpha + subj_re[,i]  + (beta+subj_re[,i+42])) - exp(alpha + subj_re[,i] - (beta+subj_re[,i+42]))
}

subjdiff<-t(subjdiff)

subjdiff<-as.data.frame(subjdiff)
colnames(subjdiff)<-c(1:42)
mns <- colMeans(subjdiff)
subjdiff<-subjdiff[,order(mns)]
```

# Example 1: Reading time differences in subject vs object relatives in English
## Individual effects of OR processing cost


```{r eval=FALSE}
for(i in 1:42){
  hist(subjdiff[,i],xlim=c(min(subjdiff),
                           max(subjdiff)),
       main=paste("subj",colnames(subjdiff)[i],sep=" "))
  Sys.sleep(.2)
}
```

# Example 1: Reading time differences in subject vs object relatives in English
## Individual effects of OR processing cost

This illustrates a point that @blastland2014norm make: ``The average is an abstraction. The reality is variation.'' 

# Example 2: Question-response accuracies (Logistic regression)

The @grodner data also has question-response accuracies: 1 if the response to a question following the sentence was correct, 0 otherwise. We show only the relevant columns below:

```{r}
head(df_gg05_rc[,c(1,2,3,4,6)])
```

# Example 2: Question-response accuracies (Logistic regression)

One could aggregate the accuracy by item, and then just fit a hierarchical linear model:

```{r}
meanp<-with(df_gg05_rc,tapply(qcorrect,
                     IND=list(condition,subj),
                     mean))
q_df<-data.frame(subj=rep(c(1:42),2),
           so=rep(c(1,-1),each=42),
           p=c(meanp[1,],meanp[2,]))

head(q_df)
```

# Example 2: Question-response accuracies (Logistic regression)

```{r}
library(lme4)
mqlmer<-lmer(p~so+(1|subj),q_df)
summary(mqlmer)
```


# Example 2: Question-response accuracies (Logistic regression)

Think about the generative process; a 0,1 response is best seen as generated by a Bernoulli distribution with probability of success $p$: $\hbox{response} \sim Bernoulli(p)$. This is the same as a Binomial process, with one trial.

One can therefore model each 0,1 response as being generated from a Bernoulli distribution, which is just a Binomial with a single trial. Thus, what is of interest is the probability of correct responses in subject vs object relatives:

```{r}
round(100*with(df_gg05_rc,
               tapply(qcorrect,condition,mean)))
```

# Example 2: Question-response accuracies (Logistic regression)

We will transform the probability $p$ of a correct response to a log-odds:

\begin{equation}
\log \frac{p}{1-p}
\end{equation}

and assume that the log-odds of a correct response is affected by the relative clause type:

\begin{equation}
\log \frac{p}{1-p} = \alpha + \beta * so 
\end{equation}

# Example 2: Question-response accuracies (Logistic regression)

This model is called a *logistic* regression because it uses the logistic or logit function to transform $p$ to log odds space. Notice that there is no residual term in this model. 

# Example 2: Question-response accuracies (Logistic regression)

We can fit the above model easily using brms:

```{r results="hide",cache=TRUE}
m_gg_q1<-brm(qcorrect~so,df_gg05_rc,
             family=bernoulli(link="logit"))
summary(m_gg_q1)
```

# Example 2: Question-response accuracies (Logistic regression)

Obviously, because the question-response data are also repeated measures, we must use a hierarchical linear model, with varying intercepts and slopes for subject and item, as in Example 1:

```{r results="hide",cache=TRUE}
m_gg_q2<-brm(qcorrect~so+(1+so|subj) + (1+so|item),
             df_gg05_rc,
             family=bernoulli(link="logit"))
summary(m_gg_q2)
```

This model is not especially good because many of the response accuracies are at ceiling. However, in principle this kind of model is appropriate for binary responses.

# Example 2: Question-response accuracies (Logistic regression)
## Convert posteriors back to probability space

What is theoretically important is the posterior distribution of the difference between object and subject relative response accuracy. That is on the probability scale.
We can go from log-odds space to probability space by solving this equation for $p$. 

Using simple algebra, we can go from:

\begin{equation}
\log \frac{p}{1-p} = \alpha + \beta * so = \mu
\end{equation}

to:

\begin{equation}
p = \exp(\mu)/(1+\exp(\mu))
\end{equation}

# Example 2: Question-response accuracies (Logistic regression)
## Convert posteriors back to probability space

For object gap sentences, the factor $so$ is coded as 1, so we have $\mu=\alpha+\beta$. For subject gap sentences, $so$ is coded as -1, so we have $\mu=\alpha-\beta$. Therefore, we just need to plug in the expression for $\mu$ for object and subject relatives.

We can now straightforwardly plot the posterior distribution of the difference between object and subject relatives. We see that there isn't any important difference between the two relative clause types. 

# Example 2: Question-response accuracies (Logistic regression)
## Convert posteriors back to probability space

```{r message=FALSE,warning=FALSE,fig.cap="\\label{fig:ggprob}The difference in question-response accuracies between object and subject relatives."}
postq<-posterior_samples(m_gg_q2)
alpha<-postq$b_Intercept
beta<-postq$b_so
mu_or<-alpha+beta
probor<-exp(mu_or)/(1+exp(mu_or))
mu_sr<-alpha-beta
probsr<-exp(mu_sr)/(1+exp(mu_sr))
```

# Example 2: Question-response accuracies (Logistic regression)
## Convert posteriors back to probability space


```{r fig.height=3}
hist(probor-probsr,freq=FALSE)
abline(v=0,lwd=2)
```

# References
